---
title: "REU 2020 Cleaned Code"
author: "James Pack"
date: "8/13/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning = F)
```

## Introduction

This code is the result of working with the University of Wisconsin at La Crosse and the USGS for a summer 2020 REU (Research Experience for Undergraduates). This group is concerned primarily with characterizing and understanding phytoplankton communities of the Upper Mississippi River System by analyzing data provided by the USGS and UMRR (Upper Mississippi River Restoration program). 

Code in this document was created by myself (James Pack). The other document with cleaned code was created by Lamia Benyamine, and the combined code and results were used in presentation to the USGS, USACE, and others. Work on a manuscript to submit for publication is ongoing.


## Research

Little research has been conducted on the phytoplankton of the UMRS. As such, we begin with a larger question and break the work into three pieces. 

"What are the main factors shaping the phytoplankton communities of the UMRS?"

After forming this question, we were able to create three shorter questions that will come together to answer our main research question. They are:

1. "How does phytoplankton biovolume and diversity vary across time and space?"
2. "What environmental and physical drivers characterize the six pools in the data?"
3. "How are phytoplankton taxa affected by changes in the environment?"

Each of the three smaller questions were tackled with slightly different approaches, and so will be answered with different chunks of code below. Brief interpretations of results as well as references to Lamia's code will be included in this document. Ideally this will ensure future work will be a direct continuation of the work compiled in the two documents. 


## Data Cleaning

The first step for all of my analyses was to load all libraries to be used and to clean and arrange the data for easy analysis later. For this, very little was changed week to week and instead the same code was reused with small additions as new variables were calculated and used in analysis. 

```{r loading libraries}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(ggthemes)
library(dplyr)
library(RColorBrewer)
library(CCA)
library(CCP)
library(vegan)
```

```{r loading data}
phytos_9606 <- read_csv('Phytos_1996-2006.csv')
phytos_0812 <- read_csv('Phytos_2008-2012.csv')
phyto_info <- read_csv('Phyto_info_WQdata.csv')

# Combining data from all time periods
phytos_whole <- phytos_9606 %>% bind_rows(phytos_0812)

# Adding water quality data, joined by SHEETBAR (ID) of sample
phyto <- phytos_whole %>% left_join(phyto_info,by="SHEETBAR")

# Cleaning the dates
phyto <- phyto %>% mutate(DATE=mdy(DATE))
phyto_info <- phyto_info %>% mutate(DATE=mdy(DATE))
```

After the data was loaded, metrics for the community were loaded to prepare for analysis. 

```{r calculating metrics}
# Richness
phyto <- left_join(phyto,phyto %>%
                     count(SHEETBAR),by='SHEETBAR') %>% 
  rename(tot_rich=n)
# Total Abundance
total <- phyto %>% group_by(SHEETBAR) %>% 
  summarize(total_abundance = sum(`TOTAL BV_um3L`))
phyto <- left_join(phyto,total,by='SHEETBAR')

# N:P ratio
np_ratio <- phyto_info %>% select(c(SHEETBAR,TN,TP)) %>% na.omit() %>% 
  mutate(np_ratio = TN/TP) %>%  # four observations missing here
  select(c(SHEETBAR,np_ratio))
phyto <- phyto %>% left_join(np_ratio,by='SHEETBAR')

## Taxa Abundance
# Cyanobacteria Abundance
cyan_abundance <- phyto %>% filter(DIVISION == 'Cyanobacteria') %>% 
  group_by(SHEETBAR) %>% 
  summarize(cyan_abund = sum(`TOTAL BV_um3L`))
phyto <- left_join(phyto,cyan_abundance,by='SHEETBAR')
phyto <- phyto %>% mutate(case_when(is.na(cyan_abund) == T ~ 0))
phyto <- phyto %>% mutate(rel_cyan = cyan_abund/total_abundance * 100)

# Bacillariophyta Abundance
bacill_abundance <- phyto %>% 
  filter(DIVISION == 'Bacillariophyta') %>% 
  group_by(SHEETBAR) %>% 
  summarize(bacill_abund = sum(`TOTAL BV_um3L`))
phyto <- left_join(phyto,bacill_abundance,by='SHEETBAR')
phyto <- phyto %>% mutate(case_when(is.na(bacill_abund) == T ~ 0))
phyto <- phyto %>% mutate(rel_bacill = bacill_abund/total_abundance * 100)

# Chlorophyta Abundance
chlor_abundance <- phyto %>% 
  filter(DIVISION == 'Chlorophyta') %>% 
  group_by(SHEETBAR) %>% 
  summarize(chlor_abund = sum(`TOTAL BV_um3L`)) 
phyto <- left_join(phyto,chlor_abundance,by='SHEETBAR')
phyto <- phyto %>% mutate(case_when(is.na(chlor_abund) == T ~ 0))
phyto <- phyto %>% mutate(rel_chloro = chlor_abund/total_abundance * 100)

# Cryptophyta Abundance
crypt_abundance <- phyto %>% 
  filter(DIVISION == 'Cryptophyta') %>% 
  group_by(SHEETBAR) %>% 
  summarize(crypt_abund = sum(`TOTAL BV_um3L`))
phyto <- left_join(phyto,crypt_abundance,by='SHEETBAR')
phyto <- phyto %>% mutate(case_when(is.na(crypt_abund) == T ~ 0))
phyto <- phyto %>% mutate(rel_crypt = crypt_abund/total_abundance * 100)

# Pyrrophyta Abundance
pyrr_abundance <- phyto %>% 
  filter(DIVISION == 'Pyrrophyta') %>% 
  group_by(SHEETBAR) %>% 
  summarize(pyrr_abund = sum(`TOTAL BV_um3L`))
phyto <- left_join(phyto,pyrr_abundance,by='SHEETBAR')
phyto <- phyto %>% mutate(case_when(is.na(pyrr_abund) == T ~ 0))
phyto <- phyto %>% mutate(rel_pyrr = pyrr_abund/total_abundance * 100)

# Chrysophyta Abundance
chrys_abundance <- phyto %>% 
  filter(DIVISION == 'Chrysophyta') %>% 
  group_by(SHEETBAR) %>% 
  summarize(chrys_abund = sum(`TOTAL BV_um3L`))
phyto <- left_join(phyto,chrys_abundance,by='SHEETBAR')
phyto <- phyto %>% mutate(case_when(is.na(chrys_abund) == T ~ 0))
phyto <- phyto %>% mutate(rel_chrys = chrys_abund/total_abundance * 100)

# Euglenophyta Abundance
eug_abundance <- phyto %>% 
  filter(DIVISION == 'Euglenophyta') %>% 
  group_by(SHEETBAR) %>% 
  summarize(eug_abund = sum(`TOTAL BV_um3L`))
phyto <- left_join(phyto,eug_abundance,by='SHEETBAR')
phyto <- phyto %>% mutate(case_when(is.na(eug_abund) == T ~ 0))
phyto <- phyto %>% mutate(rel_eug = eug_abund/total_abundance * 100)

# Creating a table of only abundances
abundances <- phyto %>% 
  select(SHEETBAR,total_abundance,
         cyan_abund,bacill_abund,chlor_abund,crypt_abund,
         pyrr_abund,chrys_abund,eug_abund)
abundances$cyan_abund <- abundances$cyan_abund %>% replace_na(0)
abundances$bacill_abund <- abundances$bacill_abund %>% replace_na(0)
abundances$chlor_abund <- abundances$chlor_abund %>% replace_na(0)
abundances$crypt_abund <- abundances$crypt_abund %>% replace_na(0)
abundances$pyrr_abund <- abundances$pyrr_abund %>% replace_na(0)
abundances$chrys_abund <- abundances$chrys_abund %>% replace_na(0)
abundances$eug_abund <- abundances$eug_abund %>% replace_na(0)
abundances <- abundances %>% 
  group_by(SHEETBAR) %>% 
  summarize(total_abundance = unique(total_abundance),
            cyan_abund = unique(cyan_abund),
            bacill_abund = unique(bacill_abund),
            chlor_abund = unique(chlor_abund),
            crypt_abund = unique(crypt_abund),
            pyrr_abund = unique(pyrr_abund),
            chrys_abund = unique(chrys_abund),
            eug_abund = unique(eug_abund))

# Simpson's D Index of Diversity
phyto <- phyto %>% group_by(SHEETBAR) %>% 
  mutate(rel_abund = `TOTAL BV_um3L`/total_abundance) %>% 
  mutate(simps_d = 1-sum(rel_abund ^ 2))
```


## Phytoplankton Communities

Our first research question focuses on the phytoplankton communities themselves. For this portion of the analysis, Lamia created most of the graphics. These graphics will be included in her document in the repository entitled "____" in the section "____". Below are the additions I had for this level of analysis, as well as what can be gathered from these analyses.

```{r biovolume trends}
phyto %>% ggplot(aes(DATE,total_abundance/1e9)) +
  geom_point() +
  geom_smooth(method=loess) +
  scale_y_log10() +
  ggtitle('Biovolume Over Time') +
  xlab('Date') +
  ylab('Biovolume (mm^3/L)') +
  theme_bw()
```

Biovolume remains mostly constant when looked at on a log scale. Some low points were observed in the later years but otherwise biovolume remained somewhat constant on the log scale. 

```{r diversity trends}
phyto %>% filter(is.na(FLDNUM) == F) %>% 
  ggplot(aes(DATE,simps_d)) +
  geom_point() +
  geom_smooth(method = loess) +
  ggtitle("Diversity Over Time") +
  xlab('Date') +
  ylab("Simpson's D Index Score") +
  theme_bw() +
  theme(axis.title=element_text())
```

An important aspect of the phytoplankton communities is the diversity of species. Communities made up of many species are healthier and more sustainable, while a community dominated by fewer species is less stable. We’ve used Simpson’s D Index as a measure of diversity for the phytoplankton communities, with scores near 1 representing the most diverse and scores near 0 representing the least diverse communities. All six stations had a wide range of scores, with the widest and narrowest ranges differing only by  0.15. Using a loess smoother, there seemed to be an upward trend in the later years, but in those same years the range of scores also increased with the lowest scores observed in 2012.

For more specific breakdowns of trends in the communities, refer to Lamia's document as it has much more details on the difference in community composition both over time and across stations. In addition to a qualitative analysis of composition, Lamia has also conducted NMDS and ANOSIM to better understand the differences in community composition.

## Characterizing the Environment

Characterizing the environment was the second objective we set for ourselves in the research endeavor. Understanding what made the six stations unique allows us to better understand what might cause the differences in community composition. For this, we chose to use CAP (Canonical Analysis of Principal Coordinates) over PCA (Principal Component Analysis) so that we could choose a non-Euclidean distance metric. With this advantage, we chose the Mahalanobis distance in order to minimize the impact outliers may have on the results. The Mahalanobis distance uses a point and a distribution to judge distance instead of two points. This allows for observations to be grouped into a distribution and a similar distance created for all points, not allowing for outliers to produce outliers in scores of the analysis.

```{r CAP}
water_data <- phyto_info %>% 
  select(SHEETBAR,FLDNUM,TN,TP,DO,SI,TEMP,TURB) %>% 
  mutate(np_ratio = TN/TP) %>% 
  select(SHEETBAR,FLDNUM,DO,SI,TEMP,TURB,np_ratio)

abund_info <- phyto %>% 
  select(SHEETBAR,total_abundance,rel_cyan,rel_bacill,rel_chloro,
         rel_crypt,rel_pyrr,rel_chrys,rel_eug) %>% 
  group_by(SHEETBAR) %>% 
  summarize(total_abund = unique(total_abundance),
            cyan_abund = unique(rel_cyan), 
            bacill_abund = unique(rel_bacill), 
            chlor_abund = unique(rel_chloro),
            crypt_abund = unique(rel_crypt),
            pyrr_abund = unique(rel_pyrr),
            chrys_abund = unique(rel_chrys),
            eug_abund = unique(rel_eug))

# fixing the NAs everywhere
abund_info[3:9][is.na(abund_info[3:9])] <- 0

community <- water_data %>%
  left_join(abund_info,by='SHEETBAR') %>% 
  na.omit() # 176 observations intact, so 4 missing


water_quality <- water_data[,-c(1,2)] %>% scale() %>% as_tibble() %>% na.omit()
species <- community[6:14] %>% scale()
community.cap <- capscale(species ~ .,
                            data=water_quality,
                            distance='mahalanobis')
plot(community.cap)
community.cap$CCA$biplot
summary(community.cap)

# plotting it better 
# I want the arrows generic plot had had
vector_points <- community.cap$CCA$biplot %>% as_tibble()
do_vector <- vector_points[1,1:2]
si_vector <- vector_points[2,1:2]
temp_vector <- vector_points[3,1:2]
turb_vector <- vector_points[4,1:2]
np_vector <- vector_points[5,1:2]

scale_vector <- .55
x <- c(vector_points[1,1], si_vector[1,1], temp_vector[1,1], 
       turb_vector[1,1], np_vector[1,1]) %>% as.numeric() * scale_vector
y <- c(vector_points[1,2], si_vector[1,2], temp_vector[1,2], 
       turb_vector[1,2], np_vector[1,2]) %>% as.numeric() * scale_vector
coords <- cbind(x,y) %>% data.frame()

my_colors <- c("Orange", "Goldenrod", "Yellow", "Blue","Violet","Navy")

community.cap$CCA$biplot %>% print()

as_tibble(community.cap$CCA$wa) %>% ggplot(aes(CAP1,CAP2)) +
  geom_jitter(aes(color=ordered_stations)) +
  xlim(-.75,.25) +
  ylim(-.5,.75) +
  scale_color_manual(values=my_colors) +
  theme_bw() +
  geom_segment(x=0,y=0,xend=coords$x[1],yend=coords$y[1],
               alpha=.5,size=.1,
               arrow=arrow(angle=25,length=unit(.1,'inches'))) +
  geom_label(x=coords$x[1]+.025,y=coords$y[1]+.06,label='DO') +
  geom_segment(x=0,y=0,xend=coords$x[2],yend=coords$y[2],
               alpha=.5,size=.1,
               arrow=arrow(angle=25,length=unit(.1,'inches'))) +
  geom_label(x=coords$x[2]+.06,y=coords$y[2],label='SI') +
  geom_segment(x=0,y=0,xend=coords$x[3],yend=coords$y[3],
               alpha=.5,size=.1,
               arrow=arrow(angle=25,length=unit(.1,'inches'))) +
  geom_label(x=coords$x[3],y=coords$y[3]-.06,label='TEMP') +
  geom_segment(x=0,y=0,xend=coords$x[4],yend=coords$y[4],
               alpha=.5,size=.1,
               arrow=arrow(angle=25,length=unit(.1,'inches'))) +
  geom_label(x=coords$x[4]-.1,y=coords$y[4],label='TURB') +
  geom_segment(x=0,y=0,xend=coords$x[5],yend=coords$y[5],
               alpha=.5,size=.1,
               arrow=arrow(angle=25,length=unit(.1,'inches'))) +
  geom_label(x=coords$x[5],y=coords$y[5]+.06,label='N:P') +
  ggtitle('CAP') +
  xlab('CAP1: Low Turbidity (38%)') +
  ylab('CAP2: High N:P Ratio (38%)') +
  labs(color='Pool')
```

There is a separation between the upper (orange/yellow) and lower (blue/purple) stations when characterizing the environmental factors. N:P ratio, dissolved oxygen, and turbidity were most important, with temperature and silica playing a lesser role in separating the stations. 






