---
title: "REU 2020 Cleaned Code"
author: "James Pack"
date: "8/13/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, warning = F)
```

## Introduction

This code is the result of working with the University of Wisconsin at La Crosse and the USGS for a summer 2020 REU (Research Experience for Undergraduates). This group is concerned primarily with characterizing and understanding phytoplankton communities of the Upper Mississippi River System by analyzing data provided by the USGS and UMRR (Upper Mississippi River Restoration program). 

Code in this document was created by myself (James Pack). The other document with cleaned code was created by Lamia Benyamine, and the combined code and results were used in presentation to the USGS, USACE, and others. Work on a manuscript to submit for publication is ongoing.


## Research

Little research has been conducted on the phytoplankton of the UMRS. As such, we begin with a larger question and break the work into three pieces. 

"What are the main factors shaping the phytoplankton communities of the UMRS?"

After forming this question, we were able to create three shorter questions that will come together to answer our main research question. They are:

1. "How does phytoplankton biovolume and diversity vary across time and space?"
2. "What environmental and physical drivers characterize the six pools in the data?"
3. "How are phytoplankton taxa affected by changes in the environment?"

Each of the three smaller questions were tackled with slightly different approaches, and so will be answered with different chunks of code below. Brief interpretations of results as well as references to Lamia's code will be included in this document. Ideally this will ensure future work will be a direct continuation of the work compiled in the two documents. 


## Data Cleaning

The first step for all of my analyses was to load all libraries to be used and to clean and arrange the data for easy analysis later. For this, very little was changed week to week and instead the same code was reused with small additions as new variables were calculated and used in analysis. 

```{r loading libraries}
library(tidyverse)
library(lubridate)
library(ggplot2)
library(ggthemes)
library(dplyr)
library(RColorBrewer)
library(CCA)
library(CCP)
library(vegan)
```

```{r loading data}
phytos_9606 <- read_csv('Phytos_1996-2006.csv')
phytos_0812 <- read_csv('Phytos_2008-2012.csv')
phyto_info <- read_csv('Phyto_info_WQdata.csv')

# Combining data from all time periods
phytos_whole <- phytos_9606 %>% bind_rows(phytos_0812)

# Adding water quality data, joined by SHEETBAR (ID) of sample
phyto <- phytos_whole %>% left_join(phyto_info,by="SHEETBAR")

# Cleaning the dates
phyto <- phyto %>% mutate(DATE=mdy(DATE))
phyto_info <- phyto_info %>% mutate(DATE=mdy(DATE))
```

After the data was loaded, metrics for the community were loaded to prepare for analysis. 

```{r calculating metrics}
# Richness
phyto <- left_join(phyto,phyto %>%
                     count(SHEETBAR),by='SHEETBAR') %>% 
  rename(tot_rich=n)
# Total Abundance
total <- phyto %>% group_by(SHEETBAR) %>% 
  summarize(total_abundance = sum(`TOTAL BV_um3L`))
phyto <- left_join(phyto,total,by='SHEETBAR')

# N:P ratio
np_ratio <- phyto_info %>% select(c(SHEETBAR,TN,TP)) %>% na.omit() %>% 
  mutate(np_ratio = TN/TP) %>%  # four observations missing here
  select(c(SHEETBAR,np_ratio))
phyto <- phyto %>% left_join(np_ratio,by='SHEETBAR')

## Taxa Abundance
# Cyanobacteria Abundance
cyan_abundance <- phyto %>% filter(DIVISION == 'Cyanobacteria') %>% 
  group_by(SHEETBAR) %>% 
  summarize(cyan_abund = sum(`TOTAL BV_um3L`))
phyto <- left_join(phyto,cyan_abundance,by='SHEETBAR')
phyto <- phyto %>% mutate(case_when(is.na(cyan_abund) == T ~ 0))
phyto <- phyto %>% mutate(rel_cyan = cyan_abund/total_abundance * 100)

# Bacillariophyta Abundance
bacill_abundance <- phyto %>% 
  filter(DIVISION == 'Bacillariophyta') %>% 
  group_by(SHEETBAR) %>% 
  summarize(bacill_abund = sum(`TOTAL BV_um3L`))
phyto <- left_join(phyto,bacill_abundance,by='SHEETBAR')
phyto <- phyto %>% mutate(case_when(is.na(bacill_abund) == T ~ 0))
phyto <- phyto %>% mutate(rel_bacill = bacill_abund/total_abundance * 100)

# Chlorophyta Abundance
chlor_abundance <- phyto %>% 
  filter(DIVISION == 'Chlorophyta') %>% 
  group_by(SHEETBAR) %>% 
  summarize(chlor_abund = sum(`TOTAL BV_um3L`)) 
phyto <- left_join(phyto,chlor_abundance,by='SHEETBAR')
phyto <- phyto %>% mutate(case_when(is.na(chlor_abund) == T ~ 0))
phyto <- phyto %>% mutate(rel_chloro = chlor_abund/total_abundance * 100)

# Cryptophyta Abundance
crypt_abundance <- phyto %>% 
  filter(DIVISION == 'Cryptophyta') %>% 
  group_by(SHEETBAR) %>% 
  summarize(crypt_abund = sum(`TOTAL BV_um3L`))
phyto <- left_join(phyto,crypt_abundance,by='SHEETBAR')
phyto <- phyto %>% mutate(case_when(is.na(crypt_abund) == T ~ 0))
phyto <- phyto %>% mutate(rel_crypt = crypt_abund/total_abundance * 100)

# Pyrrophyta Abundance
pyrr_abundance <- phyto %>% 
  filter(DIVISION == 'Pyrrophyta') %>% 
  group_by(SHEETBAR) %>% 
  summarize(pyrr_abund = sum(`TOTAL BV_um3L`))
phyto <- left_join(phyto,pyrr_abundance,by='SHEETBAR')
phyto <- phyto %>% mutate(case_when(is.na(pyrr_abund) == T ~ 0))
phyto <- phyto %>% mutate(rel_pyrr = pyrr_abund/total_abundance * 100)

# Chrysophyta Abundance
chrys_abundance <- phyto %>% 
  filter(DIVISION == 'Chrysophyta') %>% 
  group_by(SHEETBAR) %>% 
  summarize(chrys_abund = sum(`TOTAL BV_um3L`))
phyto <- left_join(phyto,chrys_abundance,by='SHEETBAR')
phyto <- phyto %>% mutate(case_when(is.na(chrys_abund) == T ~ 0))
phyto <- phyto %>% mutate(rel_chrys = chrys_abund/total_abundance * 100)

# Euglenophyta Abundance
eug_abundance <- phyto %>% 
  filter(DIVISION == 'Euglenophyta') %>% 
  group_by(SHEETBAR) %>% 
  summarize(eug_abund = sum(`TOTAL BV_um3L`))
phyto <- left_join(phyto,eug_abundance,by='SHEETBAR')
phyto <- phyto %>% mutate(case_when(is.na(eug_abund) == T ~ 0))
phyto <- phyto %>% mutate(rel_eug = eug_abund/total_abundance * 100)

# Creating a table of only abundances
abundances <- phyto %>% 
  select(SHEETBAR,total_abundance,
         cyan_abund,bacill_abund,chlor_abund,crypt_abund,
         pyrr_abund,chrys_abund,eug_abund)
abundances$cyan_abund <- abundances$cyan_abund %>% replace_na(0)
abundances$bacill_abund <- abundances$bacill_abund %>% replace_na(0)
abundances$chlor_abund <- abundances$chlor_abund %>% replace_na(0)
abundances$crypt_abund <- abundances$crypt_abund %>% replace_na(0)
abundances$pyrr_abund <- abundances$pyrr_abund %>% replace_na(0)
abundances$chrys_abund <- abundances$chrys_abund %>% replace_na(0)
abundances$eug_abund <- abundances$eug_abund %>% replace_na(0)
abundances <- abundances %>% 
  group_by(SHEETBAR) %>% 
  summarize(total_abundance = unique(total_abundance),
            cyan_abund = unique(cyan_abund),
            bacill_abund = unique(bacill_abund),
            chlor_abund = unique(chlor_abund),
            crypt_abund = unique(crypt_abund),
            pyrr_abund = unique(pyrr_abund),
            chrys_abund = unique(chrys_abund),
            eug_abund = unique(eug_abund))

# Simpson's D Index of Diversity
phyto <- phyto %>% group_by(SHEETBAR) %>% 
  mutate(rel_abund = `TOTAL BV_um3L`/total_abundance) %>% 
  mutate(simps_d = 1-sum(rel_abund ^ 2))
```


## Phytoplankton Communities

Our first research question focuses on the phytoplankton communities themselves. For this portion of the analysis, Lamia created most of the graphics. These graphics will be included in her document in the repository entitled "____" in the section "____". Below are the additions I had for this level of analysis, as well as what can be gathered from these analyses.

```{r biovolume trends}
phyto %>% ggplot(aes(DATE,total_abundance/1e9)) +
  geom_point() +
  geom_smooth(method=loess) +
  scale_y_log10() +
  ggtitle('Biovolume Over Time') +
  xlab('Date') +
  ylab('Biovolume (mm^3/L)') +
  theme_bw()
```

Biovolume remains mostly constant when looked at on a log scale. Some low points were observed in the later years but otherwise biovolume remained somewhat constant on the log scale. 

```{r diversity trends}
phyto %>% filter(is.na(FLDNUM) == F) %>% 
  ggplot(aes(DATE,simps_d)) +
  geom_point() +
  geom_smooth(method = loess) +
  ggtitle("Diversity Over Time") +
  xlab('Date') +
  ylab("Simpson's D Index Score") +
  theme_bw() +
  theme(axis.title=element_text())
```

An important aspect of the phytoplankton communities is the diversity of species. Communities made up of many species are healthier and more sustainable, while a community dominated by fewer species is less stable. We’ve used Simpson’s D Index as a measure of diversity for the phytoplankton communities, with scores near 1 representing the most diverse and scores near 0 representing the least diverse communities. All six stations had a wide range of scores, with the widest and narrowest ranges differing only by  0.15. Using a loess smoother, there seemed to be an upward trend in the later years, but in those same years the range of scores also increased with the lowest scores observed in 2012.

For more specific breakdowns of trends in the communities, refer to Lamia's document as it has much more details on the difference in community composition both over time and across stations. In addition to a qualitative analysis of composition, Lamia has also conducted NMDS and ANOSIM to better understand the differences in community composition.

## Characterizing the Environment

Characterizing the environment was the second objective we set for ourselves in the research endeavor. Understanding what made the six stations unique allows us to better understand what might cause the differences in community composition. For this, we chose to use CAP (Canonical Analysis of Principal Coordinates) over PCA (Principal Component Analysis) so that we could choose a non-Euclidean distance metric. With this advantage, we chose the Mahalanobis distance in order to minimize the impact outliers may have on the results. The Mahalanobis distance uses a point and a distribution to judge distance instead of two points. This allows for observations to be grouped into a distribution and a similar distance created for all points, not allowing for outliers to produce outliers in scores of the analysis.

```{r CAP}
water_data <- phyto_info %>% 
  select(SHEETBAR,FLDNUM,TN,TP,DO,SI,TEMP,TURB) %>% 
  mutate(np_ratio = TN/TP) %>% 
  select(SHEETBAR,FLDNUM,DO,SI,TEMP,TURB,np_ratio)

abund_info <- phyto %>% 
  select(SHEETBAR,total_abundance,rel_cyan,rel_bacill,rel_chloro,
         rel_crypt,rel_pyrr,rel_chrys,rel_eug) %>% 
  group_by(SHEETBAR) %>% 
  summarize(total_abund = unique(total_abundance),
            cyan_abund = unique(rel_cyan), 
            bacill_abund = unique(rel_bacill), 
            chlor_abund = unique(rel_chloro),
            crypt_abund = unique(rel_crypt),
            pyrr_abund = unique(rel_pyrr),
            chrys_abund = unique(rel_chrys),
            eug_abund = unique(rel_eug))

# fixing the NAs everywhere
abund_info[3:9][is.na(abund_info[3:9])] <- 0

community <- water_data %>%
  left_join(abund_info,by='SHEETBAR') %>% 
  na.omit() # 176 observations intact, so 4 missing


water_quality <- water_data[,-c(1,2)] %>% scale() %>% as_tibble() %>% na.omit()
species <- community[6:14] %>% scale()
community.cap <- capscale(species ~ .,
                            data=water_quality,
                            distance='mahalanobis')
plot(community.cap)
community.cap$CCA$biplot
summary(community.cap)

# plotting it better 
# I want the arrows generic plot had had
vector_points <- community.cap$CCA$biplot %>% as_tibble()
do_vector <- vector_points[1,1:2]
si_vector <- vector_points[2,1:2]
temp_vector <- vector_points[3,1:2]
turb_vector <- vector_points[4,1:2]
np_vector <- vector_points[5,1:2]

scale_vector <- .55
x <- c(vector_points[1,1], si_vector[1,1], temp_vector[1,1], 
       turb_vector[1,1], np_vector[1,1]) %>% as.numeric() * scale_vector
y <- c(vector_points[1,2], si_vector[1,2], temp_vector[1,2], 
       turb_vector[1,2], np_vector[1,2]) %>% as.numeric() * scale_vector
coords <- cbind(x,y) %>% data.frame()

my_colors <- c("Orange", "Goldenrod", "Yellow", "Blue","Violet","Navy")

community.cap$CCA$biplot %>% print()

as_tibble(community.cap$CCA$wa) %>% ggplot(aes(CAP1,CAP2)) +
  geom_jitter(aes(color=ordered_stations)) +
  xlim(-.75,.25) +
  ylim(-.5,.75) +
  scale_color_manual(values=my_colors) +
  theme_bw() +
  geom_segment(x=0,y=0,xend=coords$x[1],yend=coords$y[1],
               alpha=.5,size=.1,
               arrow=arrow(angle=25,length=unit(.1,'inches'))) +
  geom_label(x=coords$x[1]+.025,y=coords$y[1]+.06,label='DO') +
  geom_segment(x=0,y=0,xend=coords$x[2],yend=coords$y[2],
               alpha=.5,size=.1,
               arrow=arrow(angle=25,length=unit(.1,'inches'))) +
  geom_label(x=coords$x[2]+.06,y=coords$y[2],label='SI') +
  geom_segment(x=0,y=0,xend=coords$x[3],yend=coords$y[3],
               alpha=.5,size=.1,
               arrow=arrow(angle=25,length=unit(.1,'inches'))) +
  geom_label(x=coords$x[3],y=coords$y[3]-.06,label='TEMP') +
  geom_segment(x=0,y=0,xend=coords$x[4],yend=coords$y[4],
               alpha=.5,size=.1,
               arrow=arrow(angle=25,length=unit(.1,'inches'))) +
  geom_label(x=coords$x[4]-.1,y=coords$y[4],label='TURB') +
  geom_segment(x=0,y=0,xend=coords$x[5],yend=coords$y[5],
               alpha=.5,size=.1,
               arrow=arrow(angle=25,length=unit(.1,'inches'))) +
  geom_label(x=coords$x[5],y=coords$y[5]+.06,label='N:P') +
  ggtitle('CAP') +
  xlab('CAP1: Low Turbidity (38%)') +
  ylab('CAP2: High N:P Ratio (38%)') +
  labs(color='Pool')
```

There is a separation between the upper (orange/yellow) and lower (blue/purple) stations when characterizing the environmental factors. N:P ratio, dissolved oxygen, and turbidity were most important, with temperature and silica playing a lesser role in separating the stations. These important factors can now be considered when we move to our third question.


## Phytoplankton and the Environment

Our final research objective was to better understand the relationship between phytoplankton communities of the UMRS and their environments. Our previous work using CAP allowed us to identify the significant environmental factors: turbidity, temperature, silica, dissolved oxygen, and the nitrogen to phosphorous ratio. Knowing this, we turned to Canonical Correlation Analysis (CCA) to understand how these characterizing factors affected the phytoplankton communities.

CCA constructs canonical dimensions and then allows for the correlations between a set of X variables (the environmental factors in our case) and a set of Y variables (the phytoplankton taxa biovolume). This allows for significant dimensions to be created and then to look at the loadings of the dimension to see what effect the environment has on the phytoplankton communities. This is similar to a regression model, but allows for multiple predictor variables and multiple response variables. For this research, we were able to see how specific phytoplankton taxa react to different environmental conditions. 

To conduct this analysis, I constructed a different dataset from the existing ones in order to more easily isolate the variables of interest.

```{r preparing CCA}
# Phytoplankton Taxa Abundance
abundances <- phyto %>% 
  select(SHEETBAR,total_abundance,
         cyan_abund,bacill_abund,chlor_abund,crypt_abund,
         pyrr_abund,chrys_abund,eug_abund)
abundances$cyan_abund <- abundances$cyan_abund %>% replace_na(0)
abundances$bacill_abund <- abundances$bacill_abund %>% replace_na(0)
abundances$chlor_abund <- abundances$chlor_abund %>% replace_na(0)
abundances$crypt_abund <- abundances$crypt_abund %>% replace_na(0)
abundances$pyrr_abund <- abundances$pyrr_abund %>% replace_na(0)
abundances$chrys_abund <- abundances$chrys_abund %>% replace_na(0)
abundances$eug_abund <- abundances$eug_abund %>% replace_na(0)
abundances <- abundances %>% 
  group_by(SHEETBAR) %>% 
  summarize(total_abundance = unique(total_abundance),
            cyan_abund = unique(cyan_abund),
            bacill_abund = unique(bacill_abund),
            chlor_abund = unique(chlor_abund),
            crypt_abund = unique(crypt_abund),
            pyrr_abund = unique(pyrr_abund),
            chrys_abund = unique(chrys_abund),
            eug_abund = unique(eug_abund))

# Environmental Factors
environment <- phyto_info %>% 
  select(SHEETBAR,FLDNUM,TN,TP,DO,SI,TEMP,TURB) %>% 
  mutate(np_ratio = TN/TP) %>% 
  select(SHEETBAR,FLDNUM,DO,SI,TEMP,TURB,np_ratio) %>% 
  na.omit() # missing 4 observations

# Joining the data
community <- environment %>% 
  group_by(SHEETBAR) %>% 
  left_join(abundances %>%
              group_by(SHEETBAR),by='SHEETBAR')
community$SHEETBAR <- as_factor(community$SHEETBAR)

# Initial Correlations & X/Y setup
explanatory <- community %>% 
  group_by(SHEETBAR) %>% 
  select(TURB,np_ratio,DO,TEMP,SI)
response <- community %>% 
  group_by(SHEETBAR) %>% 
  select(bacill_abund,cyan_abund,chlor_abund,crypt_abund,
         pyrr_abund)

matcor(explanatory[,-c(1)],response[,-c(1)])$XYcor[6:10,1:5]
img.matcor(matcor(explanatory[,-c(1)],response[,-c(1)]),type=1)
```

After preparing the data and looking at the different initial correlations, CCA can be conducted to better understand the relationships in the data.

```{r CCA analysis, interpretation, and plotting}
# CCA
# thanks to https://stats.idre.ucla.edu/r/dae/canonical-correlation-analysis/
cc1 <- cc(explanatory[,-c(1)],response[,-c(1)])

# Checking the significance
rho <- cc1$cor
n <- dim(explanatory[,-1])[1]
p <- length(explanatory[,-1])
q <- length(response[,-1])

p.asym(rho, n, p, q,tstat='Wilks')

# Standardizing environmental coefficients
s1 <- diag(sqrt(diag(cov(explanatory[,-1]))))
s1 %*% cc1$xcoef

# Standardizing phytoplankton coefficients
s2 <- diag(sqrt(diag(cov(response[,-1]))))
s2 %*% cc1$ycoef

# Plotting the X/Y results for 2 significant dimensions
ordered_stations <- factor(environment$FLDNUM)
my_colors <- c("Orange", "Goldenrod", "Yellow", "Blue","Violet","Navy")

cca_x_scores <- cc1[5]$scores$xscores %>% 
  as_tibble()
cca_x_scores %>% 
  ggplot(aes(V1,V2)) +
  geom_jitter(aes(color=ordered_stations)) +
  scale_color_manual(values=my_colors) +
  theme_clean() +
  ggtitle('CCA Scores of Environmental Conditions') +
  xlab('DO & N:P Ratio') +
  ylab('Temperature') + 
  labs(color='Pool')
cca_y_scores <- cc1[5]$scores$yscores %>% 
  as_tibble()
cca_y_scores %>% 
  ggplot(aes(V1,V2)) +
  geom_jitter(aes(color=ordered_stations)) +
  scale_color_manual(values=my_colors) +
  theme_clean() +
  ggtitle('CCA Scores of Community Composition') +
  xlab('Diatom Abundance') +
  ylab('Cyanobacteria Abundance') + 
  labs(color='Pool')

img.matcor(matcor(explanatory[,-c(1)],response[,-c(1)]),type=1)
```


Variable 1:
  X: High DO, high N:P, low TURB, low TEMP, low SI
  Y: High diatom, moderately high others
  XY: Same as X
  YY: Same as Y
Variable 2:
  X: High TEMP, high SI, moderately low others
  Y: Low diatoms, high cyano, moderately high others
  XY: Same as X
  YY: Same as Y
Variable 3:
  X: Low TEMP, high SI, high TURB, high N:P,moderately high DO
  Y: Low chlorophyta, everything else negligible
  XY: Same as X
  YY: Low chlorophyta, high cryptophyta
  
Variates 3 through 5 are insignificant at the alpha = .05 level, so only variates 1 and 2 are significant for further consideration. 

First loading again emphasizes DO but now also emphasizes TEMP. Second loading emphasizes DO, TEMP, and SI, with a negative loading for N:P. 

Once again, first loading emphasizes diatoms even when standardized, second loading emphasizes a rise in cyanobacteria and drop in diatoms. Other biovolumes are minimally related to loadings when standardized. 


## Script from Presentation

In addition to all analyses and notes from the summer gathered above, below I have copied the script used for my portion of the slides that narrates the story we have attempted to tell with our analyses.


RESEARCH QUESTIONS

What are the main factors shaping the phytoplankton communities of the UMR?
1. communities
2. environment
3. communities & environment


DATA

In order to conduct our study of phytoplankton in the Upper Mississippi River System, we have worked with data collected by the USGS and the Upper Mississippi River Restoration program…. (you know the rest)

METHODS 

To better understand phytoplankton communities in the UMR, we’ve focused on our three research questions questions: What phytoplankton are being observed, what does their environment look like, and how are the phytoplankton related to their environment? Each of these questions requires a slightly different approach that will help us to better understand changes in phytoplankton communities.

We first examined the relative abundance of phytoplankton divisions in order to better understand the changes within communities. By visualizing the data, we were able to identify trends in the data across both space and time. We then used Nonmetric Multidimensional Scaling, or NMDS, to better understand how the composition of the communities differed station to station. In addition, we used ANOSIM, or analysis of similarities, to make sure that the trends were significant.

Next, to make sense of the environmental conditions at each station, we used Canonical Analysis of Principal Coordinates, or CAP, to see which factors best differentiated the six stations. This, combined with what we learned from the examination of the communities, let us know which environmental factors could be most influential in shaping the phytoplankton communities.

Finally, after identifying the significant environmental factors that varied across stations, we used Canonical Correlation Analysis, or CCA, to see how these factors were correlated with the divisions of phytoplankton observed. This final step allows us to understand what environmental factors are shaping the phytoplankton communities.


DIVERSITY I

An important aspect of the phytoplankton communities is the diversity of species. Communities made up of many species are healthier and more sustainable, while a community dominated by fewer species is less stable. We’ve used Simpson’s D Index as a measure of diversity for the phytoplankton communities, with scores near 1 representing the most diverse and scores near 0 representing the least diverse communities. All six stations had a wide range of scores, with the widest and narrowest ranges differing only by  0.15. Using a loess smoother, there seemed to be an upward trend in the later years, but in those same years the range of scores also increased with the lowest scores observed in 2012. Our next step is to look into how each station differs in its phytoplankton community composition.


CAP

Our second research objective was to better understand how the environmental factors vary across the six field stations. Because the stations were spread across hundreds of miles, the differences in phytoplankton community composition are in part explained by the varying environmental conditions at these stations. To characterize the conditions, the 181 samples were analyzed with CAP to see which factors differentiated the stations. We chose the Mahalanobis distance metric because it takes into account the variance of the factors, lessening the impact of the outliers noticed in turbidity.

The graph on the right show the results of this analysis. Along the X axis we have a component defined by low turbidity, high dissolved oxygen, and a high nitrogen to phosphorus ratio explaining 38% of the variance in environmental data. On the Y axis another 38% of the variance is explained by a high nitrogen to phosphorus ratio, high dissolved oxygen levels, and low temperatures. By coloring each point according to its location, a pattern emerges. The oranges and yellow represent the upper stations on the right side of the graph, while the purples and pink represent the lower stations on the left side of the graph. This separation seems to depend on turbidity, with a smaller emphasis on the nitrogen to phosphorus ratio and dissolved oxygen levels. The separation between the upper and lower stations also suggests that the joining of the Illinois and Mississippi River at station 4 plays some part in this distinction. Our next step is to see how the five factors seen in this graph are related to the phytoplankton communities.



CCA

Finally, we want to know how the phytoplankton communities interact with their environment. For this, we used CCA, a dimension reduction technique, to understand which factors were most influential in shaping phytoplankton communities. Five canonical dimensions were constructed, and the first two were significant. The first dimension was similar to the separation in our CAP analysis with high dissolved oxygen and nitrogen to phosphorus ratio defining the environmental variables. Overall, the community increased in biovolume, but diatoms thrived more than other phytoplankton taxa. The second dimension emphasized temperature and silica concentrations. For this dimension, there was a decrease in diatoms and an increase in cyanobacteria.However, turbidity was not an influential factor in these two dimensions. While it was a characterizing factor for the environmental conditions, it was less important in the relationships with phytoplankton communities.

Comparing these results to the initial correlation heat map shown on the right, it seems that diatoms have much stronger responses to changes in the environment than other divisions of phytoplankton. 

The bottom left corner shows the correlation between phytoplankton taxa and the environmental factors we considered.

This top row shows the correlations for diatoms specifically. The reds and blues are at either end of the spectrum, showing stronger positive and negative correlations respectively.

The second row shows the correlations for cyanobacteria. Compared to the diatoms, these correlations are more neutral, represented by the shades of green. This helps to explain why cyanobacteria increase when diatoms decrease: although the temperature may not be strongly correlated with cyanobacteria abundance, the loss of diatoms allows for other phytoplankton to flourish in their place. Similar conclusions might also be drawn about the nitrogen to phosphorus ratio and the dissolved oxygen levels; when diatoms are met with unfavorable conditions, other phytoplankton are able to flourish. 

The previous importance of turbidity, the column furthest to the left, can also contribute as it negatively affects all phytoplankton, but other conditions in the environment still allow for diatoms or cyanobacteria or other phytoplankton taxa to grow in relative abundance. 





